---
title: "Misc"
author: "Jakub Nowosad"
date: last-modified
date-format: "[Last Updated:] MMMM, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    fig-column: page
    fig-width: 10 
    fig-height: 6.5
    embed-resources: true
    toc: true
knitr:
  opts_chunk: 
    warnings: false
execute: 
  warning: false
---

```{r misc-1 }
#| message: false
#| warning: false
spain = sf::read_sf("data/spain.gpkg")
covariates = terra::rast("data/predictors.tif")
temperature = sf::read_sf("data/temp_train.gpkg")

temperature = terra::extract(covariates, temperature, bind = TRUE) |> 
  sf::st_as_sf()
```

# RandomForestsGLS

The **RandomForestsGLS** (<https://doi.org/10.21105/joss.03780>) package implements [the Generalised Least Square (GLS) based Random Forest (RF-GLS) algorithm](https://doi.org/10.1080/01621459.2021.1950003).^[Quoting the authors "RF-GLS extends RF in the same way generalized least squares (GLS) fundamentally extends ordinary
least squares (OLS) to accommodate for dependence in linear models."]
This approach is designed for spatial data modeling as it accounts for spatial dependencies in the data by:

1. Using a global dependency-adjusted split criterion and node representatives instead of the classification and regression tree (CART) criterion used in standard RF models
2. Applying contrast resampling rather than the bootstrap method used in a standard RF model
3. Employing residual kriging with covariance modeled using a Gaussian process framework

The package provides four functions:

1. `RFGLS_estimate_spatial()` for estimation in spatial data
2. `RFGLS_predict()` for prediction of the mean function
3. `RFGLS_predict_spatial()` for prediction of the spatial response
4. `RFGLS_estimate_timeseries()` for estimation in time series data (not discussed here)

The package has rather unintuitive syntax and requires the data to be in a specific format. 
We need to provide the coordinates of the data (a matrix), the response variable (a vector), and the covariates (a matrix).
In the example below, I limited the covariate matrix to the variables that are not spatial proxies.

```{r misc-2 }
library(RandomForestsGLS)

coords = sf::st_coordinates(temperature)
temp_response = temperature$temp

temperature_df = sf::st_drop_geometry(temperature)
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 7)]
covariate_matrix = as.matrix(temperature_df[, covariate_names])
```

For the example, we also split the data into training and testing sets based on created random indices.

```{r misc-3}
set.seed(2025-01-30)
train_idx = sample(1:nrow(coords), floor(nrow(coords) * 0.7))
```

The `RFGLS_estimate_spatial()` function is used to fit the RF-GLS model.
Here, we customize the number of trees to 100, but the function has many other parameters that can be adjusted.

```{r misc-4}
estimation_result = RFGLS_estimate_spatial(
  coords = coords[train_idx, ],
  y = temp_response[train_idx],
  X = covariate_matrix[train_idx, ],
  ntree = 100
)
str(estimation_result)
```

The result is a list with seven elements: a matrix of zero-indexed resamples, a matrix of predictions (ntest x ntree), a vector of predicted values, the covariate matrix, the response variable, the coordinates matrix, and and the RF-GLS object.

Now, we can use the fitted model to predict the mean function (`RFGLS_predict()`) or the spatial response (`RFGLS_predict_spatial()`).
The difference (as long as I understand) is that the former returns the mean prediction, while the latter uses the spatial coordinates in addition to the covariates to predict the spatial response.

The first function returns a list with two elements: a matrix of predictions (ntest x ntree) and a vector of predicted values, while the second function returns a list with just one element: a vector of predicted values.
Just note that the predictions by the `RFGLS_predict()` are named `"predicted"` and the predictions by the `RFGLS_predict_spatial()` are named `"prediction"`.

```{r misc-5}
#| warnings: false
prediction_result = RFGLS_predict(
  RFGLS_out = estimation_result,
  Xtest = covariate_matrix[-train_idx, ]
)
prediction_result_spatial = RFGLS_predict_spatial(
  RFGLS_out = estimation_result,
  coords.0 = coords[-train_idx, ],
  Xtest = covariate_matrix[-train_idx, ]
)
plot(prediction_result$predicted, prediction_result_spatial$prediction)
```

The final results of these two approaches are v. similar, but not identical.

Now, let's predict the models' result on the whole dataset.

```{r misc-6}
covariate_coords_r = terra::crds(covariates)
covariate_matrix_r = as.matrix(covariates)
covariate_matrix_r = covariate_matrix_r[, covariate_names]

pred_s = RFGLS_predict_spatial(
  RFGLS_out = estimation_result,
  coords.0 = covariate_coords_r,
  Xtest = covariate_matrix_r
)

pred_r = terra::setValues(covariates[[1]], pred_s$prediction)
names(pred_r) = "prediction"
terra::plot(pred_r)
```

# spatialRF

The **spatialRF** (<https://blasbenito.github.io/spatialRF/>) package aim is to provide a minimal code interface to fit spatial regression models with Random Forest.
The internal calculations are based on three general methods to generate spatial predictors from the distance matrix of the data points: Distance matrix columns as explanatory variables (Hengl et al. 2018), Moran’s Eigenvector Maps (Dray, Legendre, and Peres-Neto 2006) and PCAs.
The **ranger** package is used here internally to fit the Random Forest model.

This package also requires the data to be in a specific format.
<!-- UPDATE!!!! -->
We need to provide the data as a data frame with the dependent variable, including spatial coordinates, and the distance matrix: a matrix with the distances among the records in the data frame.

```{r misc-7 }
library(spatialRF)
library(sf)

coordinates = st_coordinates(temperature)
colnames(coordinates) = c("x", "y")
coordinates = as.data.frame(coordinates)

temperature_df = st_drop_geometry(temperature)
temperature_df$x = coordinates[, 1]
temperature_df$y = coordinates[, 2]

distance_matrix = as.matrix(dist(temperature_df[2:(ncol(temperature_df) - 9)]))
```

We also need to define the dependent variable and the predictor variables.

```{r misc-8}
response_name = "temp"
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 9)]
```

Finally, we can fit the models using one of the methods provided by the package.
The package has 10 methods implemented, nine of which are based on the three components:^[See `?rf_spatial` for more details. Also, the 10th method is `"hengl"` directly following the approach by Hengl et al. (2018).]

1. The method to generate spatial predictors (`"hengl"`, `"mem"`, or `"pca"`)
2. The method to rank spatial predictors (`"moran"`  or `"effect"`)
3. The method to select spatial predictors (`"sequential"` or `"recursive"`)

The main function of this package is `rf_spatial()`, which fits the Random Forest model with spatial predictors.
Here, an example using the Moran's Eigenvector Maps method to generate spatial predictors, Moran's I to rank them, and sequential selection of the predictors is shown.
```{r misc-9}
rf_spatial_moran = rf_spatial(
  data = temperature_df,
  dependent.variable.name = response_name,
  predictor.variable.names = covariate_names,
  distance.matrix = distance_matrix,
  distance.thresholds = 0,
  method = "mem.moran.sequential",
  n.cores = 1
)
rf_spatial_moran
```

The `rf_spatial()` returns a **ranger** model with several new slows, most importantly `residuals` that various information about the residuals, and `spatial` that contains information about the selected spatial predictors and the method used to select them.
Printing the model object provides a summary of the model, including its parameters, model performance, information on model residuals, and variable importance. 

The **spatialRF** package also provides a set of additional functions.
It includes function for reducing multicollinearity in the predictors and removing redundant spatial predictors (`filter_spatial_predictors()`); or finding promising variable interactions (`the_feature_engineer()`):

```{r misc-10}
interactions = the_feature_engineer(
  data = temperature_df,
  dependent.variable.name = response_name,
  predictor.variable.names = covariate_names,
  xy = coordinates,
  importance.threshold = 0.50, #uses 50% best predictors
  cor.threshold = 0.60, #max corr between interactions and predictors
  seed = 2025-01-30,
  repetitions = 100,
  verbose = TRUE
)
```

The `rf_evaluate()` function allows for evaluating the model using spatial cross-validation.

```{r misc-11}
rf_eval = rf_evaluate(
  model = rf_spatial_moran,
  xy = coordinates,
  repetitions = 30,
  training.fraction = 0.75,
  metrics = "rmse", 
  seed = 2025-01-30,
  verbose = TRUE
)
rf_eval
```

The `rf_importance()` function allows for visualizing the variable importance of the model.
```{r misc-12}
rf_imp = rf_importance(
  rf_spatial_moran,
  xy = coordinates
)
rf_imp
```

The `mem()` function generates Moran Eigenvector Maps (MEM) from a distance matrix.^[`mem_multithreshold()` function allows for generating MEMs for multiple distance thresholds.]

```{r misc-13}
mem1 = mem(distance.matrix = distance_matrix)
```

The package also contains a set of custom plot functions. 
One example is the `plot_response_curves()` function that allows for visualizing the response curves of the model.

```{r misc-14}
plot_response_curves(rf_spatial_moran)
```

Additional interesting functions allow for tuning the model parameters (`rf_tuning()`) or comparing several models (`rf_compare()`).
A complete list of this package's functions is available at <https://blasbenito.github.io/spatialRF/reference/index.html>.

The final prediction can be made using the `predict()` function from the **terra** package.

```{r misc-15}
pred_srf = terra::predict(covariates, rf_spatial_moran)
terra::plot(pred_srf[[1]])
```

# sperrorest

The **sperrorest** (<https://doi.org/10.32614/CRAN.package.sperrorest>) package is designed for spatial error estimation and variable importance assessment for predictive models.
The package itself does not fit the models but provides a set of functions for spatial cross-validation, including data partitioning and model cross-validation.

While the **sperrorest** package has many functions (including a set of functions for data partitioning), its main function is `sperrorest()`.
It performs spatial cross-validation for spatial prediction models, including variable importance assessment and prediction error estimation.
To use this function, we need to provide the formula, the data, the coordinates, the model function, the model arguments, the prediction function, the sampling function, and the sampling arguments.

Let's do it step by step.
First, we need to prepare the data by extracting the coordinates and creating a data frame with the dependent variable, covariates, and coordinates.

```{r misc-16 }
library(sperrorest)
library(ranger)

coordinates = sf::st_coordinates(temperature)
temperature_df = sf::st_drop_geometry(temperature)
temperature_df$x = coordinates[, 1]
temperature_df$y = coordinates[, 2]
```

Second, we need to define the formula for the model and the prediction function.

```{r misc-17}
response_name = "temp"
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 7)]
fo = as.formula(paste(response_name, "~", paste(covariate_names, collapse = " + ")))
```

Third, we need to define the custom prediction function.
The **sperrorest** package works with many model functions, but it requires a custom prediction function to extract the predictions from the model object.
In this example, we use the `ranger` model, so we need to define a custom prediction function that extracts the predictions from the `ranger` model object.
The `predict()` function from the `ranger` package returns a list with several elements, so we need to extract the predictions from this list.^[More information on the custom prediction functions is at <https://cran.r-project.org/web/packages/sperrorest/vignettes/custom-pred-and-model-functions.html>.]

```{r misc-18}
mypred = function(object, newdata) {
  predict(object, newdata)$predictions
}
```

Forth, we can perform the spatial cross-validation using the `sperrorest()` function.
We just need to provide previously prepared data, the formula, the model function, and the prediction function.
Moreover, we can also define some additional parameters of the model, such as the number of trees in the `ranger` model.
Finally, the important part is to define the sampling function (`smp_fun`) and its arguments (`smp_args`).
The sampling function is used to partition the data into training and testing sets: here we use the `partition_kmeans()` function to partition the data spatially into folds using k-means clustering of the coordinates.^[There are several other partition functions available in the package, including `partition_disc()`, `partition_tiles()`, and  `partition_cv()`.]

```{r misc-19}
# Spatial cross-validation
sp_res = sperrorest(
  formula = fo,
  data = temperature_df,
  coords = c("x", "y"),
  model_fun = ranger,
  model_args = list(num.trees = 100),
  pred_fun = mypred,
  smp_fun = partition_kmeans,
  smp_args = list(repetition = 1:2, nfold = 3),
  progress = FALSE
)
```

The result is a list with several components, including the error at the repetition and fold levels, the resampling object, the variable importance (only when `importance = TRUE`), the benchmark, and the package version.

```{r misc-20}
summary(sp_res$error_rep)
```

We can contrast the obtained results with the non-spatial cross-validation by changing the sampling function to `partition_cv()`.

```{r misc-21}
# Non-spatial cross-validation
nsp_res = sperrorest(
  formula = fo,
  data = temperature_df,
  coords = c("x", "y"),
  model_fun = ranger,
  model_args = list(num.trees = 100),
  pred_fun = mypred,
  smp_fun = partition_cv,
  smp_args = list(repetition = 1:2, nfold = 3),
  progress = FALSE
)
```

To compare both results, we can plot the RMSE values for the training and testing sets of both spatial and non-spatial cross-validation.

```{r misc-22}
library(ggplot2)
# Extract train/test RMSE from spatial CV
sp_train_rmse = sp_res$error_rep$train_rmse
sp_test_rmse = sp_res$error_rep$test_rmse
# Extract train/test RMSE from non-spatial CV
nsp_train_rmse = nsp_res$error_rep$train_rmse
nsp_test_rmse = nsp_res$error_rep$test_rmse
# Build data frame
rmse_df = data.frame(
  CV_Type = rep(c("Spatial", "Non-Spatial"), each = 4),
  Set = rep(c("Train", "Test"), each = 2),
  RMSE = c(sp_train_rmse, sp_test_rmse, nsp_train_rmse, nsp_test_rmse)
)
ggplot(rmse_df, aes(x = CV_Type, y = RMSE, fill = Set)) +
  geom_boxplot() +
  facet_wrap(~Set) +
  labs(title = "RMSE Comparison", x = "CV Method", y = "RMSE")
```

The results show that the estimation using the spatial-cross validation is less optimistic than the non-spatial cross-validation for the test set.

More examples of the package use can be found at <https://giscience-fsu.github.io/sperrorest/articles/spatial-modeling-use-case.html>/

# blockCV

The **blockCV** (<https://doi.org/10.1111/2041-210X.13107>) package provides a set of functions for block cross-validation, spatial and environmental clustering, and spatial autocorrelation estimation.
The package itself does not fit the models.

```{r misc-23 }
library(blockCV)
```

Cross-validation strategies separate the data into training and testing sets to evaluate the model's performance.
The **blockCV** package provides several cross-validation strategies, including block cross-validation, spatial clustering, environmental clustering, buffering LOO, and Nearest Neighbour Distance Matching (NNDM) LOO.

The block cross-validation is performed using the `cv_spatial()` function.
It assigns blocks to the training and testing folds randomly, systematically or in a checkerboard pattern (the `selection` argument).

```{r misc-24 }
sb1 = cv_spatial(x = temperature,
                  k = 10, # number of folds
                  size = 300000, # size of the blocks in metres
                  selection = "random", # random blocks-to-fold
                  iteration = 50, # find evenly dispersed folds
                  progress = FALSE,
                  biomod2 = TRUE)
```

The result is a list with several components, including the folds list, the folds IDs, the biomod table, the number of folds, the input size, the column name, the blocks, and the records.
For example, we can check the structure of the folds list with the `str()` function.

```{r misc-25}
str(sb1$folds_list)
```

The `cv_plot()` function additionally allows for visualizing the cross-validation results.

```{r misc-26}
cv_plot(sb1, temperature)
```

Let's compare the results of the block cross-validation with systematic and checkerboard patterns.

```{r misc-27 }
sb2 = cv_spatial(x = temperature,
                  k = 10,
                  rows_cols = c(4, 6),
                  hexagon = FALSE,
                  selection = "systematic")
cv_plot(sb2, temperature)
```

```{r misc-28 }
sb3 = cv_spatial(x = temperature,
                  k = 10,
                  size = 300000,
                  hexagon = FALSE,
                  selection = "checkerboard")
cv_plot(sb3, temperature)
```

The clustering strategies (`cv_cluster()`) are used to group the data into clusters based on spatial or environmental similarity.
The spatial similarity is based on the clustering of the spatial coordinates only. 

```{r misc-29 }
set.seed(6)
scv = cv_cluster(x = temperature, k = 10)
cv_plot(scv, temperature)
```

The environmental clustering, on the other hand, is based on the clustering of the values of the covariates extracted from the raster data.

```{r misc-30 }
set.seed(6)
ecv = cv_cluster(x = temperature, r = covariates, k = 5, scale = TRUE)
cv_plot(ecv, temperature)
```

The next cross-validation strategy is buffering LOO (also known as Spatial LOO).
It is performed using the `cv_buffer()` function, which selects a buffer around each point  (test point) and uses the points outside the buffer as the testing set.^[This approach is a form of leave-one-out cross-validation.]

```{r misc-31 }
bloo = cv_buffer(x = temperature, size = 300000, progress = FALSE)
cv_plot(bloo, temperature, num_plots = c(1, 50, 100))
```

Note that above we plot only the first, 50th, and 100th points to avoid overplotting.

The last cross-validation strategy implemented in the **blockCV** package is Nearest Neighbour Distance Matching (NNDM) LOO.
It is performed using the `cv_nndm()` function, which tries to match the nearest neighbour distance distribution function between the test and training data to the nearest neighbour distance distribution function between the target prediction and training points.
Thus, in this base, we need to provide more arguments, including a raster with the covariates, the number of samples, the sampling strategy, and the minimum training size.

```{r misc-32 }
nncv = cv_nndm(x = temperature,
                r = covariates,
                size = 300000,
                num_sample = 5000, 
                sampling = "regular",
                min_train = 0.1,
                plot = TRUE)
cv_plot(nncv, temperature, num_plots = c(1, 50, 100))
```

Let's now use the block cross-validation to fit and evaluate a model.

```{r misc-33}
# define formula
response_name = "temp"
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 7)]
fo = as.formula(paste(response_name, "~", paste(covariate_names, collapse = " + ")))

# extract the folds
folds = sb1$folds_list

model_rmse = data.frame(fold = seq_along(folds), rmse = rep(NA, length(folds)))

for(k in seq_along(folds)){
  trainSet = unlist(folds[[k]][1]) # training set indices; first element
  testSet = unlist(folds[[k]][2]) # testing set indices; second element
  rf = ranger(fo, temperature_df[trainSet, ], num.trees = 100) # model fitting on training set
  pred = predict(rf, temperature_df[testSet, ])$predictions # predict the test set
  model_rmse[k, "rmse"] = sqrt(mean((temperature_df[testSet, response_name] - pred)^2)) # calculate RMSE
}
model_rmse
```

The **blockCV** package also provides functions for checking the similarity between the folds (`cv_similarity()`) and estimating the effective range of spatial autocorrelation (`cv_spatial_autocor()`).
The first function is used to check the similarity between the folds in the cross-validation.

```{r misc-34 }
cv_similarity(cv = sb1, x = temperature, r = covariates, progress = FALSE)
```

The second function is used to estimate the effective range of spatial autocorrelation of all input raster layers or the response data -- its role is to help to determine the size of the blocks in the block cross-validation.

```{r misc-35 }
cv_spatial_autocor(r = covariates, num_sample = 5000, progress = FALSE)
```

More examples of the package's use can be found at <https://cran.r-project.org/web/packages/blockCV/vignettes/tutorial_2.html>.

# ENMeval

<https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html>

Out-of-scope: a focus on the occurrence records (ecological niche models / species distribution models)

# sits

<https://github.com/e-sensing/sits>

Out-of-scope: a focus on spatiotemporal data cubes
