---
title: "Misc"
author: "Jakub Nowosad"
date: last-modified
date-format: "[Last Updated:] MMMM, YYYY"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    fig-column: page
    fig-width: 10 
    fig-height: 6.5
    embed-resources: true
    toc: true
knitr:
  opts_chunk: 
    warnings: false
execute: 
  warning: false
---

```{r misc-1}
#| message: false
#| warning: false
spain = sf::read_sf("data/spain.gpkg")
covariates = terra::rast("data/predictors.tif")
temperature = sf::read_sf("data/temp_train.gpkg")

temperature = terra::extract(covariates, temperature, bind = TRUE) |> 
  sf::st_as_sf()
```

# RandomForestsGLS

The **RandomForestsGLS** (<https://doi.org/10.21105/joss.03780>) package implements [the Generalised Least Square (GLS) based Random Forest (RF-GLS) algorithm](https://doi.org/10.1080/01621459.2021.1950003).^[Quoting the authors "RF-GLS extends RF in the same way generalized least squares (GLS) fundamentally extends ordinary
least squares (OLS) to accommodate for dependence in linear models."]
This approach is designed for spatial data modeling as it accounts for spatial dependencies in the data by:

1. Using a global dependency-adjusted split criterion and node representatives instead of the classification and regression tree (CART) criterion used in standard RF models
2. Applying contrast resampling rather than the bootstrap method used in a standard RF model
3. Employing residual kriging with covariance modeled using a Gaussian process framework

The package provides four functions:

1. `RFGLS_estimate_spatial()` for estimation in spatial data
2. `RFGLS_predict()` for prediction of the mean function
3. `RFGLS_predict_spatial()` for prediction of the spatial response
4. `RFGLS_estimate_timeseries()` for estimation in time series data (not discussed here)

The package has rather unintuitive syntax and requires the data to be in a specific format. 
We need to provide the coordinates of the data (a matrix), the response variable (a vector), and the covariates (a matrix).
In the example below, I limited the covariate matrix to the variables that are not spatial proxies.

```{r misc-2}
library(RandomForestsGLS)

coords = sf::st_coordinates(temperature)
temp_response = temperature$temp

temperature_df = sf::st_drop_geometry(temperature)
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 7)]
covariate_matrix = as.matrix(temperature_df[, covariate_names])
```

For the example, we also split the data into training and testing sets based on created random indices.

```{r}
set.seed(2025-01-30)
train_idx = sample(1:nrow(coords), floor(nrow(coords) * 0.7))
```

The `RFGLS_estimate_spatial()` function is used to fit the RF-GLS model.
Here, we customize the number of trees to 100, but the function has many other parameters that can be adjusted.

```{r}
estimation_result = RFGLS_estimate_spatial(
  coords = coords[train_idx, ],
  y = temp_response[train_idx],
  X = covariate_matrix[train_idx, ],
  ntree = 100
)
str(estimation_result)
```

The result is a list with seven elements: a matrix of zero-indexed resamples, a matrix of predictions (ntest x ntree), a vector of predicted values, the covariate matrix, the response variable, the coordinates matrix, and and the RF-GLS object.

Now, we can use the fitted model to predict the mean function (`RFGLS_predict()`) or the spatial response (`RFGLS_predict_spatial()`).
The difference (as long as I understand) is that the former returns the mean prediction, while the latter uses the spatial coordinates in addition to the covariates to predict the spatial response.

The first function returns a list with two elements: a matrix of predictions (ntest x ntree) and a vector of predicted values, while the second function returns a list with just one element: a vector of predicted values.
Just note that the predictions by the `RFGLS_predict()` are named `"predicted"` and the predictions by the `RFGLS_predict_spatial()` are named `"prediction"`.

```{r}
#| warnings: false
prediction_result = RFGLS_predict(
  RFGLS_out = estimation_result,
  Xtest = covariate_matrix[-train_idx, ]
)
prediction_result_spatial = RFGLS_predict_spatial(
  RFGLS_out = estimation_result,
  coords.0 = coords[-train_idx, ],
  Xtest = covariate_matrix[-train_idx, ]
)
plot(prediction_result$predicted, prediction_result_spatial$prediction)
```

The final results of these two approaches are v. similar, but not identical.

Now, let's predict the models' result on the whole dataset.

```{r}
covariate_coords_r = terra::crds(covariates)
covariate_matrix_r = as.matrix(covariates)
covariate_matrix_r = covariate_matrix_r[, covariate_names]

pred_s = RFGLS_predict_spatial(
  RFGLS_out = estimation_result,
  coords.0 = covariate_coords_r,
  Xtest = covariate_matrix_r
)

pred_r = terra::setValues(covariates[[1]], pred_s$prediction)
names(pred_r) = "prediction"
terra::plot(pred_r)
```

# spatialRF

The **spatialRF** (<https://blasbenito.github.io/spatialRF/>) package aim is to provide a minimal code interface to fit spatial regression models with Random Forest.
The internal calculations are based on three general methods to generate spatial predictors from the distance matrix of the data points: Distance matrix columns as explanatory variables (Hengl et al. 2018), Moran’s Eigenvector Maps (Dray, Legendre, and Peres-Neto 2006) and PCAs.
The **ranger** package is used here internally to fit the Random Forest model.

This package also requires the data to be in a specific format.
<!-- UPDATE!!!! -->
We need to provide the data as a data frame with the dependent variable, including spatial coordinates, and the distance matrix: a matrix with the distances among the records in the data frame.

```{r misc-3}
library(spatialRF)
library(sf)

coordinates = st_coordinates(temperature)
colnames(coordinates) = c("x", "y")
coordinates = as.data.frame(coordinates)

temperature_df = st_drop_geometry(temperature)
temperature_df$x = coordinates[, 1]
temperature_df$y = coordinates[, 2]

distance_matrix = as.matrix(dist(temperature_df[2:(ncol(temperature_df) - 9)]))
```

We also need to define the dependent variable and the predictor variables.

```{r}
response_name = "temp"
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 9)]
```

Finally, we can fit the models using one of the methods provided by the package.
The package has 10 methods implemented, nine of which are based on the three components:^[See `?rf_spatial` for more details. Also, the 10th method is `"hengl"` directly following the approach by Hengl et al. (2018).]

1. The method to generate spatial predictors (`"hengl"`, `"mem"`, or `"pca"`)
2. The method to rank spatial predictors (`"moran"`  or `"effect"`)
3. The method to select spatial predictors (`"sequential"` or `"recursive"`)

The main function of this package is `rf_spatial()`, which fits the Random Forest model with spatial predictors.
Here, an example using the Moran's Eigenvector Maps method to generate spatial predictors, Moran's I to rank them, and sequential selection of the predictors is shown.
```{r}
rf_spatial_moran = rf_spatial(
  data = temperature_df,
  dependent.variable.name = response_name,
  predictor.variable.names = covariate_names,
  distance.matrix = distance_matrix,
  distance.thresholds = 0,
  method = "mem.moran.sequential",
  n.cores = 1
)
rf_spatial_moran
```

The `rf_spatial()` returns a **ranger** model with several new slows, most importantly `residuals` that various information about the residuals, and `spatial` that contains information about the selected spatial predictors and the method used to select them.
Printing the model object provides a summary of the model, including its parameters, model performance, information on model residuals, and variable importance. 

The **spatialRF** package also provides a set of additional functions.
It includes function for reducing multicollinearity in the predictors and removing redundant spatial predictors (`filter_spatial_predictors()`); or finding promising variable interactions (`the_feature_engineer()`):

```{r}
interactions = the_feature_engineer(
  data = temperature_df,
  dependent.variable.name = response_name,
  predictor.variable.names = covariate_names,
  xy = coordinates,
  importance.threshold = 0.50, #uses 50% best predictors
  cor.threshold = 0.60, #max corr between interactions and predictors
  seed = 2025-01-30,
  repetitions = 100,
  verbose = TRUE
)
```

The `rf_evaluate()` function allows for evaluating the model using spatial cross-validation.

```{r}
rf_eval = rf_evaluate(
  model = rf_spatial_moran,
  xy = coordinates,
  repetitions = 30,
  training.fraction = 0.75,
  metrics = "rmse", 
  seed = 2025-01-30,
  verbose = TRUE
)
rf_eval
```

The `rf_importance()` function allows for visualizing the variable importance of the model.
```{r}
rf_imp = rf_importance(
  rf_spatial_moran,
  xy = coordinates
)
rf_imp
```

The `mem()` function generates Moran Eigenvector Maps (MEM) from a distance matrix.^[`mem_multithreshold()` function allows for generating MEMs for multiple distance thresholds.]

```{r}
mem1 = mem(distance.matrix = distance_matrix)
```

The package also contains a set of custom plot functions. 
One example is the `plot_response_curves()` function that allows for visualizing the response curves of the model.

```{r}
plot_response_curves(rf_spatial_moran)
```

Additional interesting functions allow for tuning the model parameters (`rf_tuning()`) or comparing several models (`rf_compare()`).
A complete list of this package's functions is available at <https://blasbenito.github.io/spatialRF/reference/index.html>.

The final prediction can be made using the `predict()` function from the **terra** package.

```{r}
pred_srf = terra::predict(covariates, rf_spatial_moran)
terra::plot(pred_srf[[1]])
```

# sperrorest

```{r misc-4}
library(sperrorest)
library(rpart)

response_name = "temp"
temperature_df = sf::st_drop_geometry(temperature)
temperature_df$x = temperature_df$X
temperature_df$y = temperature_df$Y
covariate_names = colnames(temperature_df)[2:(ncol(temperature_df) - 7)]

fo = as.formula(paste(response_name, "~", paste(covariate_names, collapse = " + ")))
ctrl = rpart.control(cp = 0.005)

mypred_part = function(object, newdata) {
  predict(object, newdata)  # for regression
}

# Non-spatial cross-validation
nsp_res = sperrorest(
  data = temperature_df,
  formula = fo,
  model_fun = rpart,
  model_args = list(control = ctrl),
  pred_fun = mypred_part,
  smp_fun = partition_cv,
  smp_args = list(repetition = 1:2, nfold = 3),
  progress = TRUE
)
summary(nsp_res$error_rep)
summary(nsp_res$error_fold)

# Spatial cross-validation
sp_res = sperrorest(
  data = temperature_df,
  formula = fo,
  model_fun = rpart,
  model_args = list(control = ctrl),
  pred_fun = mypred_part,
  smp_fun = partition_kmeans,
  smp_args = list(repetition = 1:2, nfold = 3),
  progress = TRUE
)
summary(sp_res$error_rep)
summary(sp_res$error_fold)

library(ggplot2)

# Extract train/test RMSE from non-spatial CV
nsp_test_rmse = unlist(summary(nsp_res$error_rep, level = 1)$test_rmse)
nsp_train_rmse = unlist(summary(nsp_res$error_rep, level = 1)$train_rmse)

# Extract train/test RMSE from spatial CV
sp_test_rmse = unlist(summary(sp_res$error_rep, level = 1)$test_rmse)
sp_train_rmse = unlist(summary(sp_res$error_rep, level = 1)$train_rmse)

# Build data frames
nsp_df = data.frame(
  CV_Type = "Non-Spatial",
  Set = rep(c("Train", "Test"), each = length(nsp_train_rmse)),
  RMSE = c(nsp_train_rmse, nsp_test_rmse)
)
sp_df = data.frame(
  CV_Type = "Spatial",
  Set = rep(c("Train", "Test"), each = length(sp_train_rmse)),
  RMSE = c(sp_train_rmse, sp_test_rmse)
)

rmse_df = rbind(nsp_df, sp_df)

# Boxplot: training vs testing, non-spatial vs spatial
ggplot(rmse_df, aes(x = CV_Type, y = RMSE, fill = Set)) +
  geom_boxplot(position = position_dodge(0.8)) +
  theme_minimal() +
  labs(title = "RMSE Comparison", x = "CV Method", y = "RMSE")
```

# blockCV

<https://cran.r-project.org/web/packages/blockCV/vignettes/tutorial_2.html>

```{r misc-5}
library(blockCV)
```

## Block cross-validation strategies

```{r misc-6}
sb1 = cv_spatial(x = temperature,
                  k = 10, # number of folds
                  size = 300000, # size of the blocks in metres
                  selection = "random", # random blocks-to-fold
                  iteration = 50, # find evenly dispersed folds
                  progress = FALSE,
                  biomod2 = TRUE)
cv_plot(sb1, temperature)
```

```{r misc-7}
sb3 = cv_spatial(x = temperature,
                  k = 10,
                  rows_cols = c(4, 6),
                  hexagon = FALSE,
                  selection = "systematic")
cv_plot(sb3, temperature)
```

```{r misc-8}
sb4 = cv_spatial(x = temperature,
                  k = 10,
                  size = 300000,
                  hexagon = FALSE,
                  selection = "checkerboard")
cv_plot(sb4, temperature)
```

## Spatial and environmental clustering

```{r misc-9}
# spatial clustering
set.seed(6)
scv = cv_cluster(x = temperature,
                  k = 10)
cv_plot(scv, temperature)
```

```{r misc-10}
# environmental clustering
set.seed(6)
ecv = cv_cluster(x = temperature,
                  r = covariates,
                  k = 5, 
                  scale = TRUE)
cv_plot(ecv, temperature)
```

## Buffering LOO (also known as Spatial LOO)

```{r misc-11}
bloo = cv_buffer(x = temperature,
                  size = 300000, 
                 progress = FALSE)
cv_plot(bloo, temperature,
        num_plots = c(1, 50, 100))
```

## Nearest Neighbour Distance Matching (NNDM) LOO

```{r misc-12}
nncv = cv_nndm(x = temperature,
                r = covariates,
                size = 300000,
                num_sample = 5000, 
                sampling = "regular",
                min_train = 0.1,
                plot = TRUE)
cv_plot(nncv, temperature,
        num_plots = c(1, 50, 100))
```

## Check similarity

```{r misc-13}
cv_similarity(cv = ecv, # the environmental clustering
              x = temperature, 
              r = covariates, 
              progress = FALSE)
```

## Estimating size: the effective range of spatial autocorrelation

```{r misc-14}
sac1 = cv_spatial_autocor(r = covariates, 
                           num_sample = 5000, 
              progress = FALSE)
```


# ENMeval

<https://jamiemkass.github.io/ENMeval/articles/ENMeval-2.0-vignette.html>

Out-of-scope: a focus on the occurrence records (ecological niche models / species distribution models)

# sits

<https://github.com/e-sensing/sits>

Out-of-scope: a focus on spatiotemporal data cubes
