---
title: "Spatial Machine Learning with Tidymodels"
author: "Hanna Meyer"
date: last-modified
date-format: "[Last Updated:] MMMM, YYYY"
format:
  html:
    code-fold: show
    code-summary: "Hide the code"
    code-tools: true
    fig-column: page
    fig-width: 10 
    fig-height: 6.5
    embed-resources: true
    toc: true
    number-sections: true
knitr:
  opts_chunk: 
    warnings: false
execute: 
  warning: false
---

## Introduction

## Prepare data

```{r, results='hide', message=FALSE}
library(terra)
library(sf)
library(tidymodels)
library(ranger)
library(dplyr)
library(spatialsample)
library(waywiser)
library(vip)
```

Read data:

```{r, results='hide', message=FALSE}
trainingdata <- sf::st_read("data/temp_train.gpkg")
predictors <- terra::rast("data/predictors.tif")
```

Prepare data:

```{r, results='hide', message=FALSE}
trainDat <- sf::st_as_sf(terra::extract(predictors,trainingdata,bind=TRUE))
predictor_names <- names(predictors)# Extract predictor names from the raster
response_name <- "temp"
```

::: callout-note
Compared to caret, no dropping of the geometries is required
:::

## A simple model training and prediction

First, we train a random forest model.

```{r}

# Define the recipe
formula <- as.formula(paste(response_name, "~", paste(predictor_names, collapse = " + ")))
recipe <- recipes::recipe(formula, data = trainDat)

rf_model <- parsnip::rand_forest(trees = 100, mode = "regression") |>
  set_engine("ranger",importance = "impurity")

# Create the workflow
workflow <- workflows::workflow() |>
  workflows::add_recipe(recipe) |>
  workflows::add_model(rf_model)

# Fit the model
rf_fit <- parsnip::fit(workflow, data = trainDat)
```

Now let's use the model for spatial prediction

```{r}

prediction_raster <- terra::predict(predictors,rf_fit)
plot(prediction_raster)
```

## Spatial cross-validation

Define folds and the validation control

```{r}
random_folds <- rsample::vfold_cv(trainDat,v=4)
block_folds <- spatialsample::spatial_block_cv(trainDat, v = 4,n=2)
spatialsample::autoplot(block_folds)

# control cross-validation
keep_pred <- tune::control_resamples(save_pred = TRUE, save_workflow = TRUE)
```

```{r}
### Cross-validation
rf_spatial <- tune::fit_resamples(workflow, resamples=block_folds,control = keep_pred)
rf_random <- tune::fit_resamples(workflow, resamples=random_folds,control = keep_pred)

### get CV metrics
#rf_spatial$.metrics # metrics from each fold
tune::collect_metrics(rf_spatial)
tune::collect_metrics(rf_random)
```

```{r,echo=FALSE}
### analyze CV predictions
rf_spatial_pred <- data.frame(tune::collect_predictions(rf_spatial),type="spatial")
rf_random_pred <- data.frame(tune::collect_predictions(rf_random),type="random")
pred_cv <- rbind(rf_spatial_pred,rf_random_pred)
pred_cv |> 
  ggplot(aes(x = temp, y = .pred)) + 
  facet_wrap("type")+
  geom_point(alpha = .15) +
  geom_abline(color = "red") + 
  coord_obs_pred() + 
  ylab("Predicted")
```

::: callout-note
Similar to caret, we first define folds and a definition of train control. Cross-validation via fit_resamples. The final model, however, is still stored in a separate object (rf_fit)
:::

## Model tuning: spatial hyperparameter tuning and variable selection

### Hyperparameter tuning

we change the worfkflow to include the tuning specifications

```{r}
#mark two parameters for tuning:
rf_model <- parsnip::rand_forest(trees = 100, mode = "regression",
                                 mtry = tune(), min_n = tune())|>
  set_engine("ranger",importance = "impurity")


workflow <- update_model(workflow,
  rf_model)
  
  
#define tune grid:
grid_rf <- 
  grid_space_filling(        
    mtry(range = c(1, 20)), 
    min_n(range = c(2, 10)),
    size = 30) 

#tune:
rf_tuning <- tune_grid(
  workflow,
  resamples = block_folds,
  grid = grid_rf,
  control = keep_pred )
```

Analyze tuning results

```{r}
rf_tuning |> 
  collect_metrics()

rf_tuning |>
  collect_metrics() |>
  mutate(min_n = factor(min_n)) |>
  ggplot(aes(mtry, mean, color = min_n)) +
  geom_line(linewidth = 1.5, alpha = 0.6) +
  geom_point(size = 2) +
  facet_wrap(~ .metric, scales = "free", nrow = 2) +
  scale_x_log10(labels = scales::label_number()) +
  scale_color_viridis_d(option = "plasma", begin = .9, end = 0)
```

Use tuned model for prediction

```{r}
finalmodel <- fit_best(rf_tuning)
finalmodel
imp <-  extract_fit_parsnip(finalmodel) |> 
  vip::vip()
imp
terra::predict(predictors,finalmodel)
```

### Feature selection

Still missing: https://stevenpawley.github.io/recipeselectors/

## Area of Applicability, Uncertainties etc

https://applicable.tidymodels.org/ https://docs.ropensci.org/waywiser/

```{r}
model_aoa <- waywiser::ww_area_of_applicability(
  st_drop_geometry(trainDat[,predictor_names]),
  importance = vip::vi_model(finalmodel))

AOA <-  terra::predict(predictors,model_aoa)
plot(AOA$di)
plot(AOA$aoa)
```
